---
title: "Supplement 2 - Pipeline paper"
author: "Hari Rallapalli"
date: "July 23, 2018"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: united
    highlight: tango
---
```{r, echo=FALSE, warning=FALSE, message=FALSE}
rm(list = ls())
library(FSA)
library(car)
library(lme4)
library(stats)
library(dplyr)
library(ggpubr)
library(multcomp)
library(lmerTest)
```

PLX 5622 preclinical trial data analysis
===========================

This document will detail all of the methods used for statistical analyses of the PLX5622 preclinical trial dataset.

An overview of the data frame (the dataset format used in R) is provided below:

```{r, echo=FALSE, warning=FALSE}
posttreatmentpath <- file.path("Y:", "turnbulllab","turnbulllabspace","Labusers","Hari","Papers","Pipeline","Results Assets","PLX Trial Statistics - Revised JL","PLXTrial_PostreatmentVolumes.csv")
posttreatmentpath = normalizePath(posttreatmentpath)
posttreatmentdata = read.csv(posttreatmentpath)
str(posttreatmentdata)


```



Heading descriptors
----------------
- 'subject' is our animal identifier.
- 'hemisphere' specifies which side of the cerebellum, divided by the central sulcus, this specific tumor appears. Please note that some animals have two tumors, one in each hemisphere.
- 'tmxdose' specifies tumor initiation model, i.e. 1µg/g or 5 µg/g TMX
- 'sex' is the sex of the animal
- 'treatment' specifies either control (CTL) or drug (PLX) group
- 'time' is the age (in days) of the animal at the last imaging timepoint
- 'timeaftertreatment' is the time (in days) after treatment start
- 'volumecm' is the tumor volume (in cm^3) at endpoint
- 'postvolumemm' is the tumor volume (in mm^3) at endpoint
- 'prevolumemm' is the tumor volume (mm^3) at the imaging timepoint just before treatment start
- 'normalizedvolume' is equivalent to (postvolumemm-prevolumemm)/(prevolumemm*timeaftertreatment)


I am interested in the interaction between 'treatment' and 'tmxdose' as these serve as the different arms of our study.



```{r,warning=FALSE}
posttreatmentdata$cond = interaction(posttreatmentdata$treatment,posttreatmentdata$tmxdose)

posttreatmentdata %>% group_by(tmxdose,treatment) %>% summarise(n = n(), mean(prevolumemm), sd(prevolumemm), mean(postvolumemm), mean(normalizedvolume), sd(normalizedvolume))  %>% knitr::kable()


```