---
title: "PLX5622 preclinical trial statistical analyses"
author: "Hari Rallapalli"
date: "July 16, 2018"
output: 
  html_document:
    toc: true
    theme: united
---


Goals and data
==============
The goal is to iterate on the *mostly* incorrect statistical tests I was using on our preclinical drug trial tumor imaging data. I will first reproduce the incorrect results in each case. Then I will attempt to correct the results as Jason suggested.


```{r, echo=FALSE, warning=FALSE, message=FALSE}
rm(list = ls())
library(FSA)
library(car)
library(lme4)
library(stats)
library(dplyr)
library(ggpubr)
library(multcomp)
library(lmerTest)
```
Here is an overview of the data frame:
```{r, echo=FALSE, warning=FALSE}
posttreatmentpath <- file.path("Y:", "turnbulllab","turnbulllabspace","Labusers","Hari","Papers","Pipeline","Results Assets","PLX Trial Statistics - Revised JL","PLXTrial_PostreatmentVolumes.csv")
posttreatmentpath = normalizePath(posttreatmentpath)
posttreatmentdata = read.csv(posttreatmentpath)
str(posttreatmentdata)

```
Heading descriptors
----------------
- 'subject' is our animal identifier.
- 'hemisphere' specifies which side of the cerebellum, divided by the central sulcus, this specific tumor appears. Please note that some animals have two tumors, one in each hemisphere.
- 'tmxdose' specifies tumor initiation model, i.e. 1µg/g or 5 µg/g TMX
- 'sex' is the sex of the animal
- 'treatment' specifies either control (CTL) or drug (PLX) group
- 'time' is the age (in days) of the animal at the last imaging timepoint
- 'timeaftertreatment' is the time (in days) after treatment start
- 'volumecm' is the tumor volume (in cm^3) at endpoint
- 'postvolumemm' is the tumor volume (in mm^3) at endpoint
- 'prevolumemm' is the tumor volume (mm^3) at the imaging timepoint just before treatment start
- 'normalizedvolume' is equivalent to (postvolumemm-prevolumemm)/(prevolumemm*timeaftertreatment)


I am interested in the interaction between 'treatment' and 'tmxdose' as these serve as the different arms of our study.


```{r,warning=FALSE}
posttreatmentdata$cond = interaction(posttreatmentdata$treatment,posttreatmentdata$tmxdose)

posttreatmentdata %>% group_by(tmxdose,treatment) %>% summarise(n = n(), mean(prevolumemm), sd(prevolumemm), mean(postvolumemm), mean(normalizedvolume), sd(normalizedvolume))  %>% knitr::kable()


```


Pre-treatment analyses - Incorrect
----------------------
Repeating the original, however incorrect, Kruskal-Wallis Test on the pre-treatment tumor volumes and apply Dunn's multiple comparisons test without correction.


```{r}
posttreatmentdata %>% ggboxplot(x = "cond", y = "prevolumemm", add = "jitter", xlab = "condition", ylab = "pretreatment volume (mm^3)")

kruskal.test(prevolumemm ~ cond, data = posttreatmentdata)

dunnTest(prevolumemm ~ cond, data = posttreatmentdata, method = "none")


```
Pre-treatment analyses - Corrected
----------------------

 Now I need to correct the pre-treatment tumor volume analysis by accounting for the potential effect of two tumors coming from the same subject using mixed effects model. 
 
 Am I am correctly accounting for the effect of two tumors within the same subject? I used Table 2 on page 6 of Bates et. al. *Journal of Statistical Software* 2014 (https://arxiv.org/pdf/1406.5823.pdf) as a guide.
 
 Is it correct to forgo fdr multiple comparisons correction when using Tukey's method (https://www.itl.nist.gov/div898/handbook/prc/section4/prc471.htm)?

```{r}

pretreatmentmodel = lmer (prevolumemm ~ 0 + cond + (1|subject), data = posttreatmentdata) 

pretreatmentmodel %>% summary()



pretreatmentmodel %>% glht(linfct = mcp(cond="Tukey")) %>% summary()


```
My interpretation is that there is still no significant difference between the arms of our study in terms of pre-treatment tumor volume. Is this correct?

Post-treatment analyses - Incorrect
----------------------
First, visualize the data and repeat the incorrect analysis. Kruskal-Wallis of the normalized tumor volume data and Dunn's multiple comparisons without correction.

```{r}

kruskal.test(normalizedvolume ~ cond, data = posttreatmentdata)

dunnTest(normalizedvolume ~ cond, data = posttreatmentdata, method = "none")


posttreatmentdata %>% ggboxplot(x = "cond", y = "normalizedvolume", add = "jitter", xlab = "condition", ylab = "posttreatment normalized volume")
posttreatmentdata %>% ggboxplot(x = "cond", y = "postvolumemm", add = "jitter", xlab = "condition", ylab = "posttreatment volume (mm^3)")

```

Post-treatment analyses - Corrected
----------------------
I attempt to correct our post-treatment tumor volume analysis by accounting for the potential effect of two tumors coming from the same subject using a mixed effects model. 

```{r}
posttreatmentmodel = lmer (normalizedvolume ~ 0 + cond + (1|subject), data = posttreatmentdata) 

posttreatmentmodel %>% summary()

posttreatmentmodel %>% glht(linfct = mcp(cond="Tukey")) %>% summary()


```

My interpretation of these results is that the differences in normalized tumor volume between PLX and CTL groups are not significant in either model. However, the contrasts (estimates) between PLX and CTL groups in both the SMO1 and SMO5 cohorts are relatively large. Does this suggest an effect, or am I not allowed to say that using this test and/or given the small sample size?



Pre/Post-Treatment analysis - Chris' suggested model
----------------------------

If I understood his email correctly, Chris suggested adding 'prevolumemm' as a fixed effect and use (un-normalized)post-treatment tumor volume as response! He also suggested adding 'tmxdose' and 'treatment' as fixed effects in addition to their interaction. I will do both, but I will also build a model with only the 'prevolumemm' fixed effect addition just for fun(?).



Starting with the addition of 'prevolumemm' to the model as a fixed effect ONLY:
```{r}
prepostmodel = lmer(postvolumemm ~ 0 + prevolumemm  + cond + (1|subject), data = posttreatmentdata) ## why can't I pipe the data to the model? :(


prepostmodel %>% summary()
prepostmodel %>% ranova()

```

The tmxdose:treatment effects (estimates) are negative in the drug-treated condition, whereas they are positive in the control condition in either tumor production model. My interpretation is that the treatment (PLX) reduces endpoint tumor volume (postvolumemm) with respect to untreated animals (CTL), and that this effect is more apparent in tumors from the SMO5 model than from the SMO1 model. Pre-treatment volume correlates positively with post-treatment volume, as expected. Is this correct?

Chris mentioned that I could just look at the associated t-statistic to test these effects, but I am not sure how to do this without plugging in the t-statistic and df = 51 into a t-distribution table and getting a p-value (two-tailed hypothesis). Am I doing this correctly? I won't pretend to completely understand this paper, but is it correct to even be looking for a p-value (https://onlinelibrary.wiley.com/doi/epdf/10.1111/j.1467-9450.2010.00852.x)?



```
Fixed effects:
             Estimate Std. Error t value        p value
prevolumemm    1.4287     0.4273   3.344        .001554
condCTL.SMO1   1.5595    10.7354   0.145        .885282
condPLX.SMO1  -0.4145    10.9921  -0.038        .969836
condCTL.SMO5   8.0541    11.5198   0.699        .487728
condPLX.SMO5 -10.0343    12.9722  -0.774        .442504
```


Given all of this, I built the complete model as Chris suggested. I am hesitant to interpret the results as I am not able to display all of the estimates. Is the addition of 'tmxdose' and 'treatment' as fixed effects truly adding new, meaningful information over only including their interaction?


```{r,message=FALSE}
prepostmodel_full = lmer(postvolumemm ~ 0 + tmxdose + treatment + prevolumemm  + cond + (1|subject), data = posttreatmentdata) 
prepostmodel_full %>% summary() 

  
```